Modelling the creation of ecological traps by environmentally persistent pathogens

### 2015-06-04 Manuscript/manuscript.tex
Small edits and exploratory plots

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

@@

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study then, we seek to explore how a pathogen's traits, including its environmental longevity (how long it can persist and remain infectious in the environment), interact with the distribution of habitat quality to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.

@@

The model was parameterized  according to the following assumptions (Table~\ref{params}). (1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  (2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).  (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).  To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity were varied over a range of possible values.

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with either high or low quality habitat distributions (with patch qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24, respectively).  Specifically, we implemented a full-factorial design in which disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity, representing the half-life of the pathogen's infectivity in the environment, was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

@@

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating migrants that select all patches with equal probability).  To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk) and the final total effective host population size.

@@

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was determined largely by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

However, despite facilitating pathogen spread, for pathogens with either low or high disease-induced mortalities, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (Fig~\ref{highvlow}).  
In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported greater occupancy than low quality distributions due to the greater stability and connectivity afforded by high quality habitat.
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely regardless of habitat quality, high quality distributions again supported larger host populations than low quality distributions.

For pathogens falling between these extremes (i.e. with intermediate disease-induced mortality), the habitat quality distribution had a substantial effect on pathogen spread.  
In particular, for a given disease, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).  
For a given pathogen in this range (here with $\nu = 0.2$ and $\delta = 0.3$), the relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the more pronounced effect of increased longevity on pathogen persistence and spread in high quality habitat distributions.
Again, metapopulations with a lattice structure displayed similar patterns, but required lower disease-induced mortality to see the benefit of low quality habitat (Supplement).

To better understand the mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).  
Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
However, disease risk, measured by the proportion of time a patch is occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  
Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e. $\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs. d-f).
Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially compared to random habitat selection (i.e. $\xi_{em} = 0$, Fig~\ref{pext}).

>>>>

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

@@

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies, \cite{George2013}), and thus understanding how pathogen transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study then, we seek to explore how a pathogen's traits, including its environmental longevity (how long it can persist and remain infectious in the environment), interact with the distribution of habitat quality to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.

@@

The model was parameterized  according to the following assumptions (Table~\ref{params}). (1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  (2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).  (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).  To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity (representing the half-life of the pathogen's infectivity in the environment) were varied over a range of possible values.

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with either high or low quality habitat distributions (with patch qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24, respectively).  Specifically, we implemented a full-factorial design in which disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

@@

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk) and the final total effective host population size.

@@

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

Despite facilitating pathogen spread, for pathogens with either low or high disease-induced mortalities, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (Fig~\ref{highvlow}).  
In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat.
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely regardless of habitat quality, high quality distributions again supported larger host populations than low quality distributions.

For pathogens falling between these extremes (i.e. with intermediate disease-induced mortality), the habitat quality distribution had a substantial effect on pathogen spread.  
In particular, for these pathgoens, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).  
For a given pathogen in this range (here with $\nu = 0.2$ and $\delta = 0.3$), the relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the more pronounced effect of increased longevity on pathogen persistence and spread in high quality habitat distributions (note that this is not a general trend and in some cases is even reversed).
Again, metapopulations with a lattice structure displayed similar patterns, but required lower disease-induced mortality to see the benefit of low quality habitat (Supplement).

To better understand the mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).  
Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
However, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  
Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e. $\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs. d-f).
Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially compared to random habitat selection (i.e. $\xi_{em} = 0$, Fig~\ref{pext}).
